\documentclass[defaultstyle,11pt]{thesis}

\usepackage{amssymb}		% to get all AMS symbols
\usepackage{graphicx}		% to insert figures

\usepackage{color}
\usepackage{soul} % \hl \ul \st \caps \so
\usepackage{float}
\usepackage{apacite} % \cite<prefix>[suffix]{key1,key2}, \citeA, \citeNP,
                     % \fullcite, \shortcite, \citeyear, \citeauthor, \nocite
% \usepackage{changebar}
% \nochangebars

\synctex=1

\pdfinfo{
   /Author (Matthew V. Mollison)
   /Title (Dissertation)
   /Subject (EEG correlates of the spacing effect)
   /Keywords (Spacing effect, SME, EEG, ERP,
   oscillations, encoding, learning)
}
%%%%%%%%%%%%   All the preamble material:   %%%%%%%%%%%%

\title{Distributed practice and distributed representations: Investigating the spacing effect using EEG}

\author{M.\ V.}{Mollison}

\otherdegrees{B.A., Brandeis University, 2005 \\
	      M.A., University of Colorado Boulder, 2010}

\degree{Doctor of Philosophy}		%  #1 {long descr.}
	{Ph.D., Cognitive Neuroscience}		%  #2 {short descr.}

\dept{Department of}			%  #1 {designation}
	{Psychology and Neuroscience}		%  #2 {name}

\advisor{Prof.}				%  #1 {title}
	{Tim Curran}			%  #2 {name}

\reader{Prof.~Alice F. Healy}		%  2nd person to sign thesis
\readerThree{Prof.~Albert Kim}		%  3rd person to sign thesis
\readerFour{Prof.~Leaf Van Boven}		%  3rd person to sign thesis
\readerFive{Prof.~Michael C. Mozer}		%  3rd person to sign thesis

\abstract{  \OnePageChapter	% because it is very short

% Often the abstract will be long enough to require
% more than one page, in which case the macro
% ``$\backslash$OnePageChapter'' should {\it not} be used. But this one isn't, so it should.

The spacing effect shows that studying information at distributed intervals leads to better long-term memory than massing study episodes together over the same cumulative amount of time.
Prior research has not made explicit predictions about neural activity in the EEG domain that should occur at different study repetition lags under different spacing effect theories.
One main purpose of this thesis is to make these predictions in terms of the neural activity expected under different spacing effect hypotheses, while another is to test these predictions.  Knowledge about neural patterns underlying the spacing effect can shed light on why this effect occurs, which would support or challenging the hypotheses presented here as well as possibly provide a physiological grounding when considering the spacing effect in educational settings.  We used event-related potentials and time--frequency methods to analyze spaced and massed study repetitions, as well as measures of neural similarity across presentations to inform these predictions.
To summarize our findings, when studying information for a second time at a spaced interval, the retrieval of the initial study episode from long-term memory and the additional semantic processing received benefits performance on a subsequent test compared to studying massed repetitions.
% The present results show that the spacing effect is supported by the retrieval of a prior study instance, following years of behavioral results and now revealing the neural characteristics of this process.
}

%\dedication[Dedication]{	% NEVER use \OnePageChapter here.
%    Dedication?
%	}

\acknowledgements{	\OnePageChapter	% *MUST* BE ONLY ONE PAGE!
    First and foremost, thanks to my advisor, Dr.\ Tim Curran, for being a remarkable mentor who was always available to discuss ideas about research and analyses, who inspired me to pursue interesting and novel research topics, and who supported me throughout my graduate training.

    Thanks to many of the faculty associated with the Department of Psychology and Neuroscience at the University of Colorado Boulder, particularly those who I learned from and worked with over the years.  Thanks as well to Dr.\ Michael Kahana, whose lab sparked my interest in research on human memory when I was an undergraduate.

	  Thanks to Dr.\ Tim Curran, Dr.\ Alice Healy, Dr.\ Albert Kim, Dr.\ Leaf Van Boven, and Dr.\ Michael Mozer for being on my dissertation committee.

    I would like to acknowledge and thank Chris Bird and Krystin Corby, the professional research assistants who helped organize and run participants for my experiments.  I also could have not completed research nearly so quickly if it were not for the undergraduate research assistants who have been part of the Curran Lab over the years.

    Thanks to Dr.\ Jeremy Manning and Dr.\ Phillip Gilley for discussions about and assistance with EEG similarity analyses.

    Thanks to my family and friends for their valuable support over the years.  You will be pleased to know that I am excited and happy with where pursuing a graduate degree has led me.  I would not have made it here without following that path and your encouragement along the way.

    This work was supported by NSF grant SMA-1041755 to the Temporal Dynamics of Learning Center.

    %Here's where you acknowledge folks who helped.
	%But keep it short, i.e., no more than one page,
	%as required by the Grad School Specifications.
	}

\IRBprotocol{13-0462}	% optional!

% \ToCisShort	% use this only for 1-page Table of Contents

% \LoFisShort	% use this only for 1-page Table of Figures
% \emptyLoF	% use this if there is no List of Figures

% \LoTisShort	% use this only for 1-page Table of Tables
\emptyLoT	% use this if there is no List of Tables

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%       BEGIN DOCUMENT...         %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\input specific_aims.tex
\input background_sig.tex
\input dependent_measures.tex
% \input research_prop_general.tex
\input exp_1.tex
% \input research_prop_exp.tex
\input exp_2.tex
\input general_discussion.tex

%%%%%%%%%   then the Bibliography, if any   %%%%%%%%%
%\bibliographystyle{plain}   % or "siam", or "alpha", etc.
\bibliographystyle{apacite}   % or "siam", or "alpha", etc.
% \nocite{*}		% list all refs in database, cited or not
% \bibliography{refs}		% Bib database in "refs.bib"
\bibliography{/Users/matt/Library/texmf/bibtex/bib/mvm_tex/mvm}
%\bibliography{mvm}

% % Remove issue number from all citations: http://tex.stackexchange.com/questions/170070/biblatex-apa-style-shouldnt-show-issue-number-in-references-for-journal-publica
% \AtEveryBibitem{\ifentrytype{article}{\clearfield{number}}{}}

%%%%%%%%%   then the Appendices, if any   %%%%%%%%%
% \appendix
% \input appendixA.tex
% \input appendixB.tex


\end{document}

